Compositional Quantitative Resource Analysis of C Programs

### 2014-06-16 paper/main.tex
more work on section 2

To this end, we fix the shape of the potential functions $\Phi$ so
that it becomes possible to use \emph{linear programming} to compute a
derivation in the logic.  If we assume that a program state $\state$
simply maps variable to integers then we require
$$
\Phi(\state) = \sum_{x,y \in \dom{\state}} q_{(x,y)} {\cdot} |[\state(x),\state(y)]|
$$
where $q_{(x,y)} \in \Qplus$ and $|[a,b]| = \max(b-a,0)$.

\subsection{Motivating Examples}

We now present a representative set of examples that can be
handled by our tool.  The bounds that are given are for the number
of ``base statements'' executed by the program, that is, the number
of assertions, skip and assignment statements executed.  The
bounds are given as literal expressions where a variable name {\tt x}
represents the initial value of this variable.

\paragraph{Simple loops.} The simplest loop scheme that is in
use in typical imperative programs is counting upwards or
downwards to a given bound N.  The two example programs
are given below, on their right is the bound inferred by our analysis.
\begin{center}
\begin{tabular}{p{0.7\textwidth} | p{0.2\textwidth} }
\lstinline|while x > N ( x = x - 1 )| & $|{\tt N} -  {\tt x}|$ \\
\lstinline|while x < N ( x = x + 1 )| & $|{\tt N} -  {\tt x}|$ \\
\end{tabular}
\end{center}

It is also frequent to iterate every $n$ elements in an array, also
in this simple case our analysis works without any problem.  Note
however that if we increment using a variable, it might very well be
possible that the program diverges!  In this situation no worst
case bound can be given and the analysis correctly fails.
\begin{center}
\begin{tabular}{p{0.7\textwidth} | p{0.2\textwidth} }
\lstinline|while x <= -2 ( x = x + 2 )| & $| {\tt x} | / 2$ \\
\lstinline|while x <= z - y ( x = x + y )| & Infeasible \\
\lstinline|assert y > 0; while x <= z - y ( x = x + y )| & $1 + |{\tt x} - {\tt z}|$
\end{tabular}
\end{center}
In the last example, the constant 1 is added to the loop bound
to account for the initial \texttt{assert} that consumes one
step.  Also, since the bound is for the worst case, it is tight
only when {\tt y} is 1.

\paragraph{In the literature.}
The two examples below are taken from Gulwani's paper~\cite{Gulwani-speed}.
They are both handled by the Speed tool but require inference of a so-called
disjunctive invariant.  In the abstract interpretation community, these
invariants are known to be notoriously difficult to handle.

\begin{minipage}{0.45\textwidth}
\begin{lstlisting}
while x < n
  if y < m    y = y + 1
  else        x = x + 1
\end{lstlisting}
$$
|{\tt m} - {\tt y}| + |{\tt n} - {\tt x}|
$$
\end{minipage}
\begin{minipage}{0.45\textwidth}
\begin{lstlisting}
while x < n
  if z > x    x = x + 1
  else        z = z + 1
\end{lstlisting}
$$
|{\tt n} - {\tt x}| + |{\tt n} - {\tt z}|
$$
\end{minipage}

\hspace{1ex}

While the first
example is simple to understand, the second one is much more
tricky and trying to understand how it actually works is quite challenging.
However, with the amortized analysis in mind, using the potential transfer
reasonning, it is almost trivial to prove the termination.  While the Speed
tool has to find a fairly involved invariant for the loop, our tool is simply
reasoning locally and works without any special trick.
%
Our analysis handles without problems all the other examples
in~\cite{Gulwani-speed} that have a linear bound.  Here is
a quick review of some of them together with the bounds we
inferred.

\begin{minipage}[b]{0.30\textwidth}
\begin{lstlisting}
while x < n
  while x < n
    x = x + 1
\end{lstlisting}
$$
|{\tt x} - {\tt n}|
$$
\end{minipage}
\begin{minipage}[b]{0.30\textwidth}
\begin{lstlisting}
while x < n (x = x + 1);
while x < n (x = x + 1)
\end{lstlisting}
$$
|{\tt x} - {\tt n}|
$$
\end{minipage}
\begin{minipage}[b]{0.30\textwidth}
\begin{lstlisting}
while x < n
  ( while y < m (y = y+1)
  ; x = x + 1 )
\end{lstlisting}
$$
|{\tt x} - {\tt n}|
+|{\tt y} - {\tt m}|
$$
\end{minipage}

\paragraph{Amortized reasonning.}
The two examples below show the power of amortized reasoning.
Because they essentially rely on an amortized reasonning we believe
that our tool is the only one able to deal with them.

\begin{minipage}[b]{0.45\textwidth}
\begin{lstlisting}
assert y >= 1;
while x + y <= N
  ( x = x + y
  ; z = y; z = z + 4
  ; while z > 0 ( z = z - 1 ) )
\end{lstlisting}
$$
1 + 8\,|{\tt N} - {\tt x}|
$$
\end{minipage}
\begin{minipage}[b]{0.45\textwidth}
\begin{lstlisting}
while x > 0
  ( x = x - 1
  ; if ?
      while s > 0 ( s = s - 1 )
    else s = s + 1 )
\end{lstlisting}
$$
3\,|{\tt x}| + |{\tt s}|
$$
\end{minipage}

\hspace{1ex}

Maybe surprisingly, the two examples have linear bounds.  In the first one, the
worst case is when {\tt y} is 1, in this situation the inner loop will perform 5 times
$|{\tt N} - {\tt x}|$ iterations. Indeed, for each decrement of {\tt x}, {\tt z} is set
to 5 and iterated over.  The second example is classic in the amortized analysis
community, it is the original example from Tarjan's paper~\cite{Tarjan-amort} that
non deterministically pushes one element on a stack or pops many elements from
that stack.  In the example source code, the stack depth is represented by the
variable {\tt s} and the number of iterations is ruled by the initial value of {\tt x}.

@@

\appendix

@@

In this appendix we provide a non-exhaustive catalog of classes of
programs that can be automatically analyzed by our system.  For
simplicity, we use cost metric that counts cost $1$ for assignments
and cost $0$ for all other operations.  Sometimes we also use the
\emph{ticks metric} if we want to discuss features like returning
resources.  Of course, the examples can also be analyzed with any other
cost metric.

@@

$3|[0,x]| + |[0,y]|$
\\[.7\baselineskip]
      {\bf t13}
    \end{center}
  \end{minipage}
%
%
%
  \begin{minipage}[b]{\progwidth}
    \begin{center}
   \begin{lstlisting}
while (n<0) {
  n=n+1;
  y=y+1000;
  while (y>=100){
    y=y-100;
  }
}
   \end{lstlisting}

@@

$1+3|[0,x]|$
\\[.7\baselineskip]
      {\bf t15}
    \end{center}
  \end{minipage}
%
%
  \begin{minipage}[b]{\progwidth}
    \begin{center}
   \begin{lstlisting}
  assert (y>=0);
  while (x-y>0) {
    x=x-1;
    x=x-y;
    z=y;
    z=z+y;
    z=z+100;
    while (z>0) {
      z=z-1;
    }
  }
   \end{lstlisting}

@@

\pref{fig:cat1} shows code snippets that need amortization are
compositionality to obtain a whole program bound.

Example \emph{t07} demonstrates two different features of the
analysis.  For one thing it shows that we can precisely track size
changes inside loops.  In the first loop, we increment $y$ by $2$ in
each of the $|[0,x]|$ iterations.  For another thing it shows that we
automatically recognize dead code if we find conflicting assertions on
a branching path: After the second loop we know $y \leq 0$ and as a
result can assign arbitrary potential inside the third loop where we
know that $y>0$.  As a result, we obtain a tight bound.

@@

In example \emph{t10} we also do not restrict the inputs $x$ and $y$.
They can be negative, positive, or zero.  The star {\tt *} in the
conditional, stands for an arbitrary assertion.  In each branch of the
conditional we can obtain a constant potential $1$ since the interval
size $|[y,x]|$ is decreasing.

Example \emph{t13} shows how amortization can be used to handle tricky
nested loops.  The outer loop is iterated $|[0,x]|$ times.  In the
conditional, we either (the branching condition is again arbitrary)
increment the variable $y$ or we execute an inner loop in which $y$ is
counted back to $0$.  Then analysis computes a tight linear bound for
this program.  Again, the constants $0$ and $1$ in the inner loop can
as well be replace by something more interesting, say $9$ and $10$
like in example \emph{t08}.  Then we still obtain a tight linear
bound.

Example \emph{t27} is similar to example \emph{t13}.  Instead of
decrements the variable $x$ in the outer loop we this time increment
the variable $n$ till $n = 0$.  In each of the $|[n,0]|$ iterations,
we increment the variable $y$ by $1000$.  We then execute an inner
loop that increments $y$ by $100$ until $y=0$.  The analysis can
derive that only the first execution of the inner loop depends on the
initial value of $y$.  We again derive a tight bound.

Example \emph{t28} is particularly interesting.  In the first loop we
decrement the size $|[y,x]|$.  However, we also shift the interval
$[y,x]$ to the interval $[y+1000,x+1000]$.  The analysis can derive
that this does not change the size of the interval and computes the
tight loop bound $3|[y,x]|$.  The additional to loops are in the
program to show that the size tracking in the first loop works
accurately.  The second loop is executed $|[0,y]| + 1000|[y,x]|$ times
in the worst case.  The third loop is executed $|[x,0]| + |[y,x]|$ in
the worst case (if $x$ and $y$ are negative).

Sometimes we need some assumptions on the inputs in order to derive a
bound.  Example \emph{t15} is such a case.  We assume here that the
input variable $y$ is non-negative and write \code{assert(y>=0)}.  The
semantic of \code{assert} is that it has no effect if the assertion is
true and that the program is terminated without further cost
otherwise.  The cost of an assert is $1$ resource unit in our cost
metric.  If we enter the loop then we know that $x>0$ and we can
obtain constant potential from the assignment $x=x-1$.  After the
assignment we know that $x\geq y$ and $y\geq 0$.  As a consequence, we
can share the potential $3[0,x]$ before the assignment $x=x-y$ between
$3|[0,x]|$ and $3|0,y|$ after the assignment.  In this way, we derive
a tight linear bound.

Example \emph{t16} is an extension of example \emph{t15}. We again
assume that $y$ is non-negative and use the same mechanism to iterate
the outer loop as in \emph{t15}.  In the inner loop, we also count the
variable $z$ down to zero and perform $|[0,z]|$ interations.  However,
insead of assigning $z=y$, we assign $z=2y+100$.  The analysis
computes the tight linear bound $1+2|[0,x]|+103|[y,x]|$.  The
assignment of potential to the size interval $|[y,x]|$ in instead of
$|[0,x]|$ is a random choice of the LP solver.  Other optimal
solutions of the linear program that the analyzer gernerates for
\emph{t15} would result in bounds such as $1+105|[0,x]|$ or
$1+6|[0,x]|+99|[y,x]|$.  In fact, the $105$ potential units can be
shared arbitrarily between the interval sizes $|[y,x]|$ and $|[0,x]|$
as long as more than $1$ unit is assigned to $|[0,x]|$.  Since we know
that $y\geq 0$ the choice of the LP solver is optimal but this seems
to be a coincidence.

@@

Example \emph{t20} shows how we can handle programs in which bounds
contain absolute values like $|x-y|$.  The first loop body is only
executed if $x<y$ and the second loop body is only executed if $y<x$.
The analyzers finds a tight bound.

At first sight, example \emph{t30} appears to be a simple loop that
decrements the variable $x$ down to zero.  However, a closer look
reveals that the loop actually decrements both input variables $x$ and
$y$ down to zero before terminating.  In the loop body, first $x$ is
decremented by one.  Then the values of the variables $x$ and $y$ are
switched using the local variable $t$ as a buffer.  Out analysis
infers the tight bound $4|[0,x]|+4|[0,y]|$.

@@

Examples \emph{fig2\_1} and \emph{fig2\_2} are taken from Gulwani' et
al~\cite{Gulwani-speed}.  They are both handled by the SPEED tool but
require inference of a \emph{disjunctive invariant}.  In the abstract
interpretation community, these invariants are known to be notoriously
difficult to handle.
%
In example \emph{fig2\_1} we have one loop that first increments
variable $y$ up to $m$ and then increments variable $x$ up to $n$.  We
derive the tight bound $|[x, n]| + |[y, m]|$.
%
Example \emph{fig2\_2} is more tricky and trying to understand how it
works may be challenging.  However, with the amortized analysis in
mind, using the potential transfer reasoning, it is almost trivial to
prove the termination.  While the SPEED tool has to find a fairly
involved invariant for the loop, our tool is simply reasoning locally
and works without any clever tricks. We obtain the tight bound $|[x,
n]| + |[z, n]|$.

>>>>

To this end, we fix the shape of the potential functions $\Phi$ so
that it becomes possible to use \emph{linear programming} (LP) to
compute a derivation in the logic.  If we assume that a program state
$\state$ simply maps variable to integers then we require
$$
\Phi(\state) = q_0 + \sum_{x,y \in \dom{\state}} q_{(x,y)} {\cdot} |[\state(x),\state(y)]|
$$
where $q_{(x,y)} \in \Qplus$ and $|[a,b]| = \max(b-a,0)$.  Note that
$q_0$ is the constant potential and that we treat constants as program
variables.  For instance, we always have $q_{(0,x)}|[0,x]|$ as a
component of $\Phi(\state)$ if $x \in \dom{\state}$.

We then develop an inference rule for each syntactic construct that
derives sound triples $\htriple{\Phi}p{\Phi'}$ in the quantitative
logic but enables inference using and LP solver.  This idea is best
explained by example.  Assume for the remainder of this section that
we are interested in bounding the number of \emph{ticks} in a program.
In other words, we use a resource metric that assigns cost $n$ to the
function call \code{tick(n)} and cost $0$ to all other operations.
The following example can then be bounded by $|[x,y]|$.
\begin{lstlisting}
  while (x<y) {x=x+1; tick(1);}
\end{lstlisting}
To derive this bound in our automatic amortized analysis, we start
with the initial potential $\Phi_0 = |[x,y]|$ (short for
$\Phi_0(\state) = |[\state(x),\state(y)]|$) which we also use as the
loop invariant.  For the loop body we then (like in Hoare logic) have
to derive a triple like
$\htriple{\Phi_0}{\code{x=x+1;tick(1)}}{\Phi_0}$.  However, we can
only do so if we exploit the fact that $x<y$ at the beginning of the
loop body.  The reasoning then works as follows.  We start with the
potential $|[x,y]|$ and the fact that $|[x,y]| > 0$ before the
assignment.  If we denote the update version of $x$ after the
assignment by $x'$ then the relation $|[x,y]| = |[x',y]| + 1$ between
the potential before and after the assignment \code{x=x+1} holds.
This means that we have the potential $|[x,y]| + 1$ before the
statement $tick(1)$.  Since \code{tick(1)} consumes one resource unit,
we end up with potential $|[x,y]|$ after the loop body and have
established the loop invariant again.

More formally, we develop (in section XXX) a judgement 
$$
\Gamma; Q \vdash S \dashv Q'; \Gamma'
$$
such that $\Gamma$ contains logical assertions like $x<y$ that we
collect along the conditional branches of the program and $Q$ is a
family of coefficents $Q = (q_{(a,b)})_{a,b \in \text{scope}}$ that is
indexed by the variables that are currently in scope.  The logical
assertions in $\Gamma$ are of the form XXXX.  They are only used to
determine at the assignment if we can extract constant potential to
pay for future cost.  It is important to note that the reasoning the
assertions in $\Gamma$ are very basic.  We do not perform any fixpoint
computations or invariants.

Note that our notion of a potential function is a refinement of the
concept of a ranking function.  A potential function can be used like
a ranking function if we use the tick metric and add \code{tick(1)} to
every back edge (loops and function calls) of the program.  However, a
potential function is more flexible.  For example, we can use a
potential function to prove that the following program does not
consume any resources in the tick metric.
\begin{lstlisting}
  while (x<y) {tick(-1); x=x+1; tick(1);}
\end{lstlisting}
Similarly we can prove that the following program can be bounded by
$10|[x,y]|$.  In both cases, we reason exactly like in the first
version of the while loop to prove the bound.  Of course, such loops
with different notations can be seamlessly combined in a larger
program.
\begin{lstlisting}
  while (x<y) {x=x+1; tick(10);}
\end{lstlisting}

To understand how the analysis works, it is instructive to look at some
more involved examples.

@@

\cleardoublepage
\appendix

@@

In this appendix we provide a non-exhaustive catalog of classes of
programs that can be automatically analyzed by our system.  For
simplicity, we use a cost metric that counts cost $1$ for assignments
and cost $0$ for all other operations.  Sometimes we also use the
\emph{ticks metric} if we want to discuss features like returning
resources.  Of course, the examples can also be analyzed with any other
cost metric.

@@

$3|[0,x]| + |[0,y]|$
\\[.7\baselineskip]
      {\bf t13}
    \end{center}
  \end{minipage}
%
%
%
  \begin{minipage}[b]{\progwidth}
    \begin{center}
   \begin{lstlisting}
while (n<0) {
  n=n+1;
  y=y+1000;
  while (y>=100 && *){
    y=y-100;
  }
}
   \end{lstlisting}

@@

$3|[0,x]|$
\\[.7\baselineskip]
      {\bf t15}
    \end{center}
  \end{minipage}
%
%
  \begin{minipage}[b]{\progwidth}
    \begin{center}
   \begin{lstlisting}
  assert (y>=0);
  while (x-y>0) {
    x=x-1;
    x=x-y;
    z=y;
    z=z+y;
    z=z+100;
    while (z>0) {
      z=z-1;
    }
  }
   \end{lstlisting}

@@

Figures~\ref{fig:cat1a}, \ref{fig:cat1b}, and~\ref{fig:cat1c} show
code snippets that need amortization are compositionality to obtain a
whole program bound.

Example \emph{t07} demonstrates two different features of the
analysis.  For one thing it shows that we can precisely track size
changes inside loops.  In the first loop, we increment $y$ by $2$ in
each of the $|[0,x]|$ iterations.  An in the second loop, we decrement
$y$.  For another thing it shows that we automatically recognize dead
code if we find conflicting assertions on a branching path: After the
second loop we know $y \leq 0$ and as a result can assign arbitrary
potential inside the third loop where we know that $y>0$.  As a
result, we obtain a tight bound.

@@

In example \emph{t10} we also do not restrict the inputs $x$ and $y$.
They can be negative, positive, or zero.  The star {\tt *} in the
conditional, stands for an arbitrary assertion.  In each branch of the
conditional we can obtain the constant potential $1$ since the interval
size $|[y,x]|$ is decreasing.

Example \emph{t13} shows how amortization can be used to handle tricky
nested loops.  The outer loop is iterated $|[0,x]|$ times.  In the
conditional, we either (the branching condition is again arbitrary)
increment the variable $y$ or we execute an inner loop in which $y$ is
counted back to $0$.  The analysis computes a tight linear bound for
this program.  Again, the constants $0$ and $1$ in the inner loop can
as well be replace by something more interesting, say $9$ and $10$
like in example \emph{t08}.  Then we still obtain a tight linear
bound.

Example \emph{t27} is similar to example \emph{t13}.  Instead of
decrementing the variable $x$ in the outer loop we this time increment
the variable $n$ till $n = 0$.  In each of the $|[n,0]|$ iterations,
we increment the variable $y$ by $1000$.  We then execute an inner
loop that increments $y$ by $100$ until $y=0$.  The analysis can
derive that only the first execution of the inner loop depends on the
initial value of $y$.  We again derive a tight bound.

Example \emph{t28} is particularly interesting.  In the first loop we
decrement the size $|[y,x]|$.  However, we also shift the interval
$[y,x]$ to the interval $[y+1000,x+1000]$.  The analysis can derive
that this does not change the size of the interval and computes the
tight loop bound $3|[y,x]|$.  The additional two loops are in the
program to show that the size tracking in the first loop works
accurately.  The second loop is executed $|[0,y]| + 1000|[y,x]|$ times
in the worst case.  The third loop is executed $|[x,0]| + |[y,x]|$ in
the worst case (if $x$ and $y$ are negative).

Sometimes we need some assumptions on the inputs in order to derive a
bound.  Example \emph{t15} is such a case.  We assume here that the
input variable $y$ is non-negative and write \code{assert(y>=0)}.  The
semantic of \code{assert} is that it has no effect if the assertion is
true and that the program is terminated without further cost
otherwise.  If we enter the loop then we know that $x>0$ and we can
obtain constant potential from the assignment \code{x=x-1}.  After the
assignment we know that $x\geq y$ and $y\geq 0$.  As a consequence, we
can share the potential $3|[0,x]|$ before the assignment \code{x=x-y}
between $3|[0,x]|$ and $3|0,y|$ after the assignment.  In this way, we
derive a tight linear bound.

Example \emph{t16} is an extension of example \emph{t15}. We again
assume that $y$ is non-negative and use the same mechanism to iterate
the outer loop as in \emph{t15}.  In the inner loop, we also count the
variable $z$ down to zero and perform $|[0,z]|$ interations.  However,
instead of assigning \code{z=y}, we assign \code{z=2y+100}.  The analysis
computes the tight linear bound $1+2|[0,x]|+103|[y,x]|$.  The
assignment of potential to the size interval $|[y,x]|$ in instead of
$|[0,x]|$ is a random choice of the LP solver.  Other optimal
solutions of the linear program that the analyzer gernerates for
\emph{t15} would result in bounds such as $1+105|[0,x]|$ or
$1+6|[0,x]|+99|[y,x]|$.  In fact, the $105$ potential units can be
shared arbitrarily between the interval sizes $|[y,x]|$ and $|[0,x]|$
as long as more than $1$ unit is assigned to $|[0,x]|$.  Since we know
that, $y\geq 0$ the choice of the LP solver is optimal but this seems
to be a coincidence.

@@

Example \emph{t20} shows how we can handle programs in which bounds
contain absolute values like $|x-y|$.  The first loop body is only
executed if $x<y$ and the second loop body is only executed if $y<x$.
The analyzer finds a tight bound.

At first sight, example \emph{t30} appears to be a simple loop that
decrements the variable $x$ down to zero.  However, a closer look
reveals that the loop actually decrements both input variables $x$ and
$y$ down to zero before terminating.  In the loop body, first $x$ is
decremented by one.  Then the values of the variables $x$ and $y$ are
switched using the local variable $t$ as a buffer.  Our analysis
infers the tight bound $4|[0,x]|+4|[0,y]|$.

@@

Examples \emph{fig2\_1} and \emph{fig2\_2} are taken from Gulwani et
al~\cite{Gulwani-speed}.  They are both handled by the SPEED tool but
require inference of a \emph{disjunctive invariant}.  In the abstract
interpretation community, these invariants are known to be notoriously
difficult to handle.
%
In example \emph{fig2\_1} we have one loop that first increments
variable $y$ up to $m$ and then increments variable $x$ up to $n$.  We
derive the tight bound $|[x, n]| + |[y, m]|$.
%
Example \emph{fig2\_2} is more tricky and trying to understand how it
works may be challenging.  However, with the amortized analysis in
mind, using the potential transfer reasoning, it is almost trivial to
prove a bound.  While the SPEED tool has to find a fairly
involved invariant for the loop, our tool is simply reasoning locally
and works without any clever tricks. We obtain the tight bound $|[x,
n]| + |[z, n]|$.
